Incremental tactics

### 2012-05-31 doc/papers/LFMTP2012/paper.tex
paper

The positive impact of rich type systems on programming language
design is not to be demonstrated anymore, both in terms of the safety
of programs --- they ``do not go wrong'', and programming practices
--- modularity, early bug catching. Verifying well-typing of
real-world programs is thus a critical matter, but it is a complex
operation, for the compiler designer and for the human operator:

@@

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for what is commonly referred as ``type checking'',
or explicit ``church-style'' calculi. Take for example dependent type
systems, relying on a certain notion of definitional equality on
types, or type systems with subtype polymorphism: both rely on a rule
of the form
$$
\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}
$$
(for a given notion of $\leq$) which is not syntax-directed. The
implementer then has to convert the declarative system to an
algorithm, and convince himself that both are equivalent. This task is
highly non-trivial and non-modular: for instance, how to infer the
type of a System \sysname{T} recursor
$$
\infer{
  \Gamma\vdash M : \cst{nat}
  \and
  \Gamma\vdash N : A
  \and
  \Gamma, \var x:\cst{nat}, \var y:A\vdash P : A
}{
  \Gamma\vdash \recb M N x y P
}
$$
in presence of subtyping? The problem is that we have to guess a type
$A$ more general than that of $N$ and $P$, while giving that type to
variable $\var y$. A well-known algorithm if the subtyping relation
forms a complete lattice is to iterate the typing of $P$ with the join
of the previously computed and the new type until a fixpoint is
reached.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to take only the result of the algorithm into
consideration. But let us choose another path for a minute:
derivations in the declarative systems, if they cannot always be
easily inferred, can be \emph{verified} if provided by an oracle. If
the oracle is wrong, we can tell cheaply by verifying the witness
typing derivation it provides. Let us then imagine a type-checker to
be the pair of 1/ an untrusted procedure $\finfer M$ to infer a typing
derivation out of program $M$; it doesn't have to be complete, or even
sound, and 2/ a trusted and fast kernel $\fcheck{\md}$ to verify these
derivations in a given set of inference rules $\Sigma$. This is a
\emph{certifying} scheme (see \cite{leroy2006compcert} for a
discussion on approaches to reliable compilers):
if $\fcheck{\finfer M} = \cst{true}$ then there really \emph{is} a
derivation $\md$ for $M$, but the converse is not necessarily true:
function $\finfer{}\!\!$ could fail, or have bugs (return an incorrect
derivation). Explicitly generating the derivation in the declarative
system has the following advantages over the first approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item compilers often need to propagate type information further in
  the chain of transformation; the typing derivation produced by
  $\finfer{}$ being explicit, it could constitute the data structure
  being exploited.
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\end{itemize}
\ldots\ but there is more:

@@

As type systems become richer, it becomes increasingly difficult for
the programmer to write a correct program in one try, compile it and
run it: writing a program longer than a few dozens of lines becomes a
tight and non-linear interaction between the human and the
type-checker involving constant experiments, fixes, etc. If this
interaction is possible for short programs and fast checkers, it
becomes increasingly hindered by the latency of complete rechecks:
even if the modifications made to the program between two interactions
are small, the whole program is usually rechecked. This is especially
true for languages embedding formal verification aspects, like proof
assistants: there, tight interaction with a type or proof checker is
unavoidable due to the difficulty of writing correct proofs without
guidance from the system, and the time taken by many tools to infer
easy parts of the proofs (proof search). The constant modification of
the source makes it necessary to observe \emph{incrementally} the
effect of a small change on the whole edifice: we need to take
advantage of the results of previous checks in order to build the
result for a modified version of the program.

@@

Now, if we consider as above a type-checker as a derivation-generating
procedure, and storing these derivations between calls to the
procedure, we could recognize already-built subderivations for
subterms common to a checked and a new version of the program, and
graft these pieces of derivation to form a new, large one for the
whole modified program, saving the regeneration of all equal subterms'
derivations in compatible contexts. This way we constantly maintain
the verifiable witness of well-typing for the program, and take
advantage of the quotients made on higher-order syntax.

We can thus add to the previous list that explicitly generating typing
derivations makes possible a form of safe incrementality by sharing
common subderivations.

@@

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping on those types. He (or
she) wants to write a program computing $(2+2)\times 3$. He knows that
sometimes those $\cst{rec}$ arguments are quite hard to get correct,
but \sysname{T_{<:}} is a typed language; he wants to take advantage
of this to build his program incrementally using the teacher's
provided incremental verifier. As a first attempt, he wants to verify
$P_1$:
$$
\recb{\s\z}{\s\z} x y {\s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$, and
he could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \\&\qquad
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program. Derivation $\md_2$
used to live in the empty context, we are then free to use it in the
new context where $\var{add}:\cst{nat}\to\cst{nat}$, because System
\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
with the derivation $\finfer{\var u}$ inferred recursively for then
new recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{odd}$, but $\var{add}$ awaits $\cst{nat}$ arguments. When the
system will generate this derivation, it will insert a coercion
$\cst{odd} \leq \cst{nat}$ at these two points. Taking advantage of
$\md_2$ and $\md_4$ the system succeeds with a new derivation made out
of these, without having to regenerate them because it knows that
$\finfer{\fget\md} = \md$.

@@

\begin{mathleft}
  \quad\case{\app{M}N} \\
  \quad\quad\letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
  \quad\quad\letd {\pair {A_2} {\md_2}} {\finfer{N}} \\
  \quad\quad\letd {\md_{\leq}} {\fleq {A_1} {A_2}} \\
  \quad\quad
  \match{\md_{\leq}} \\
  \quad\quad\quad\case{\infer{ }{A\leq A}}
  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}} \\
  \quad\quad\quad\case{\_}
  \pair {B} {\infer{\md_1 \and \infer*{\md_2 \and \md_{\leq}}{\vdash N
      : A_1}}{\vdash \app M N : B}}
\end{mathleft}

\noindent
The $\syntax{let}$ construct is just syntactic sugar for a one-branch
$\syntax{case}$; we use an auxiliary function $$\fleq{}{}\ :\
\prd A {\cst{tp}} \prd B {\cst{tp}} \vdash A\leq B = \ldots$$ trying to
construct a subtyping derivation for its argument, or failing with an
error (definition not included here). Notice the slight optimization
to the size of the derivation: if the types $A_1$ and $A_2$ are the
same syntactically, we don't generate a subtyping rule. This is done by
matching on the returned derivation $\md_\leq$.

Then comes the abstraction case $\tlam x A M$. The difficulty is that
the recursive call $\finfer{M}$ must be done in an enlarged
environment where $\vdash x : A$, otherwise the check that the
returned open derivation $\md$ is correct would fail ($x$ is
unknown). We thus introduce a new construct $\matchin M \Gamma \ldots$
and its $\syntax{let}$ counterpart ($\match M \ldots$ being syntactic
sugar for $\matchin M \envnil \ldots$). Even then, in the recursive
call, arriving to variable $\var x$ would fail since we would have no
way of telling its type anymore. Thus we substitute all occurrences of
$\var x$ with the \emph{output} of function $\finfer{}$ on it (since
it is known at this point), \ie\ a pair of $A$ and $\md_1$. For this
substitution to be well-typed, we now need to coerce this output to a
valid input (a term). We introduce an operator which is in some way
the \emph{inverse} of $\finfer{}$: operator $\fget{}$ maps a pair of a
type and a derivation to the term it types.

@@

At this point, we get back $B$ and $\md$ from the recursive call. We
know that $B$ can't depend on $\md_x$ (because a type is always
closed), but $\md$ can: it is not \emph{per se} a hypothetical
derivation but an \emph{open} derivation where a special judgement
named $\md_x$ can occur. To return the complete derivation of
$\vdash\tlam A x M : A \to B$, we must then close, or ``hypothesize''
$\md$ by $\md_x$, and then apply rule \rulename{Lam}:

@@

For what concerns the treatment of environments, the case of $\recb M
N x y P$ is similar: we first compute recursively the derivations
$\md_M$ and $\md_N$ which should be well-typed in the current
environment, and then $\md_P$ in the enlarged environment, with
variables $\var x$ and $\var y$ in $P$ substituted with the (future)
result of $\finfer{}$. The problem is with subtyping: terms $N : A_N$
and $P : A_P$ should be coercible to the same type $A$, but $P$ should
be typed under $\vdash y:A$, which is not known before typing $P$. The
trick to break the loop is to iterate the typing of $P$ with more and
more general types, until we reach a fixpoint (this is finite if there
is a unique more general type); in our case, that means iterating
twice: we will then attain \cst{nat}.

@@

The type-checker is now done. There is no case for variables, as we
won't ever meet this case: we already substituted all variables with
their result derivation. We only need to make sure that in this case,
the derivation will be returned:
$$
\finfer{\fget{\pair A \md}} = \md
$$
Now $\fget{}$ can be derived automatically from the type of
$\finfer{}$: it is its inverse. It has definition:

\begin{mathleft}
  \fget{}\ :\ \prd{\{M\}}{\cst{tm}} \prd {A}{\cst{tp}} \prd{\md}{(\vdash
    M:A)} \cst{tm} =
  \\\quad \lamd M \lamd A \lamd \md M
\end{mathleft}

@@

A similar idea is very wide-spread in \sysname{LCF}-style proof
assistants: during manual proof search, the interaction is driven by
\emph{tactics}: %TODO cite
they are untrusted programs generating pieces of proofs, assembled by
other, higher-order tactics; verification of the generated proofs is
done when the evaluation of all tactics is complete. The differences
with our mechanism are two-fold: first, tactics usually take one (or
several) \emph{goal}(s) as input, \ie\ holes in a proof, whereas our
functions can take any term; secondly, proof-check is usually done at
the end of the evaluation, which can lead to less tractable bugs; by
making evaluation and checking mutually recursive, we choose to detect
errors earlier in these witnesses: exactly at the point where they are
generated.

Introducing the ``inverse'' of the derivation inference function has
two purpose: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special variable
construct $\{x:A\}$ denoting free variables is added to the syntax of
terms, and the rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would arrive to the unhandled case $\finfer{\var
  x}$. This requires evaluating the type-checker with a carefully
designed strategy (see \ref{sec:comput}).

@@

To write deltas by sharing subterms, we need to be able to address any
subterm and instantiate its free variables. For this, we define the
the \emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
metavariables $\meta X$, $\meta Y$\ldots\ as in
\cite{nanevski2008contextual}. Metavariables stand for well-typed
terms defined in a repository, and are instantiated wrt. their free
variables. A repository also contains a distinguished metavariable
corresponding to the head of the term stored in it. For instance, the
repository
$$
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{o}} &:
    \cst{tm}

\end{array}\right), \smeta X {}
$$
is one of the possible way of slicing of the term
$$
\fbox{$\app {\cst{lam}} \lam x
\fbox{$\app {\cst{lam}} \lam y
\fbox{\app {\cst{app}} \app {\var y}
}$}
{(\fbox{\app
  {\cst{app}}\app{\var x}{\cst o}})
}
$}$$
where all boxed subterms are referable slices.

@@

We distinguish syntactically between \emph{checkable} and
\emph{inferrable} objects (resp. \emph{canonical} and \emph{atomic}),
an idea dating back to \cite{pierce2000local}: canonical objects are
checked against a given type, whereas we can synthesize the type of
atomic objects; the coercion from the first to the second will be the
place where to verify equality between the synthesized type and the
given one.

@@

Note that this notion of substitution differs from hereditary
substitution found in \eg\ \cite{hl07mechanizing} as it presupposes
$\eta$-long normal forms: once a \emph{cut} starts, there must be
exactly as many formal and actual arguments. A non-empty spine cut
against another non-empty spine $\gcut {\app H S} {S'}$ blocks the
substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
implicit $\eta$-expansion); conversely, exhausting actual but not
formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
substitution too. For example, the reduction $\gcut {\lam x \app {\var
    x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck because the
second $\var x$ in the left part is substituted by a function yet is
not itself a function.

@@

Our algorithm types exactly the well-typed objects of the usual
\LF, except that we thread a repository $\mr$ and store in it
all well-typed applicative subterms $F$ along with their classifier
$A$ and the environment in which they are well-typed.

@@

\begin{theorem}
  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jem\mmr\Gamma M {M'} A$
\end{theorem}

\begin{theorem}
  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jlfm\Gamma {(\checkoutrr
    M)} (\checkoutrr A)$
\end{theorem}

@@

\section{Implementation}
\label{sec:implem}
\section{Related work}
% 1-2 pages

>>>>

The positive impact of rich type systems on programming language
design is not to be demonstrated anymore, both in terms of the safety
of program execution --- they ``do not go wrong'', and programming
practices --- modularity, early bug catching. Verifying well-typing of
real-world programs is thus a critical matter, but it is a complex
operation, for the compiler designer and for the human operator:

@@

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for what is commonly referred as ``type checking'',
or explicit ``church-style'' calculi. Take for example dependent type
systems, relying on a certain notion of definitional equality on
types, or type systems with subtype polymorphism: both rely on a rule
of the form
$$
\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}
$$
(for a given notion of $\leq$) which is not syntax-directed. The
implementer then has to convert the declarative system to an
algorithm, and convince himself that both are equivalent. This task is
highly non-trivial and non-modular: for instance, how to infer the
type of a System \sysname{T} recursor $\recb M N x y P$, in presence
of subtyping? The problem is that we have to guess a type $A$ more
general than that of $N$ and $P$, while giving that type to variable
$\var y$. A well-known algorithm if the subtyping relation forms a
complete lattice is to iterate the typing of $P$ with the join of the
previously computed and the new type until a fixpoint is reached.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to take only the result of the algorithm into
consideration. But let us choose another path for a minute:
derivations in the declarative systems, if they cannot always be
easily inferred, can be \emph{verified} if provided by an oracle. If
the oracle is wrong, we can tell cheaply by verifying the witness
typing derivation it provides. Let us then imagine a type-checker to
be the pair of \emph{(i)} an untrusted procedure $\finfer M$ to infer
a typing derivation out of program $M$; it doesn't have to be
complete, or even sound, and \emph{(ii)} a trusted and fast kernel
$\fcheck{\md}$ to verify these derivations \wrt\ a given set of inference
rules $\Sigma$. This is a \emph{certifying} scheme (see
\cite{leroy2006compcert} for a discussion on approaches to reliable
compilers): if $\fcheck{\finfer M} = \cst{true}$ then there really
\emph{is} a derivation $\md$ for $M$, but the converse is not
necessarily true: function $\finfer{}\!\!$ could fail, or have bugs
(return an incorrect derivation). Explicitly generating the derivation
in the declarative system has the following advantages over the first
approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item compilers often need to propagate type information further in
  the chain of transformation; the typing derivation produced by
  $\finfer{}$ being explicit, it could constitute the data structure
  being exploited.
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\end{itemize}
\ldots\ but there is more:

@@

As type systems become richer, it becomes increasingly difficult for
the programmer to write a well-typed program in one try, compile it
and run it: writing a program longer than a few dozens of lines
becomes a tight and non-linear interaction between the human and the
type-checker involving constant experiments, fixes, etc. If this
interaction is possible for short programs and fast checkers, it
becomes increasingly hindered by the latency of complete rechecks:
even if the modifications made to the program between two interactions
are small, the whole program is usually rechecked. This is especially
true for languages embedding formal verification aspects, like proof
assistants: there, tight interaction with a type or proof checker is
unavoidable due to the difficulty of writing correct proofs without
guidance from the system, and the time taken by many tools to infer
easy parts of the proofs (proof search). The constant modification of
the source makes it necessary to observe \emph{incrementally} the
effect of a small change on the whole edifice: we need to take
advantage of the results of previous checks in order to build the
result for a modified version of the program.

@@

Now, if we consider as above a type-checker as a derivation-generating
procedure, and storing these derivations between calls to the
procedure, we could recognize already-built subderivations for
subterms common to an old, checked version and a new, to-be-checked
version, and graft together these pieces of derivation to form a new,
large one for the new version of the program, saving the regeneration
of all equal subterms' derivations in compatible contexts. This way we
constantly maintain the verifiable witness of well-typing for the
program, and take advantage of the quotients made on higher-order
syntax.

We can thus add to the list of advantages begun earlier that
explicitly generating typing derivations makes possible a form of safe
incrementality by sharing common subderivations.

@@

\begin{figure}
  \centering
%  \small
  \begin{mathpar}
    \infer{ }{\jts{\cst o} {\even}}

    \infer{\jts M \odd} {\jts {\s M} \even}

    \infer{\jts M \even} {\jts {\s M} \odd}

    \infer{\jts M \nat} {\jts {\s M} \nat}

    \infer{\jts M {A\to B} \and \jts N A} {\jts {\app M N} B}

    \infer{\infer*{}{\mbox{$
          \begin{array}{c}
            [\jts {\var x} A] \\[-0.3em]
            \vdots\\
            \jts M B
          \end{array}$
        }}} {\jts {\lam x M} {A\to B}}

    \infer{\jts M A \and \jsub A B}{\jts M B}

    \infer{\jts M \nat
      \and
      \jts N A
      \and
      \infer*{}{\mbox{$
        \begin{array}{c}
          [\jts {\var x} \nat] \quad
          [\jts {\var y} A] \\[-0.3em]
          \vdots \\
          \jts P A
        \end{array}
        $}}
    }{\jts {\recb M N x y P} A}

    \infer{ }{\jsub A A}

    \infer{ }{\jsub \even\nat}

    \infer{ }{\jsub \odd\nat}

    \infer{\jsub {A'} A \and \jsub B {B'} }{\jsub {A\to B}{A'\to B'}}
  \end{mathpar}
  \caption{System \sysname{T_{<:}}, declaratively}
  \label{fig:tsub}
\end{figure}

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping on those types; see
Fig. \ref{fig:tsub}. He (or she) wants to write a program computing
$(2+2)\times 3$. He knows that sometimes those $\cst{rec}$ arguments
are quite hard to get correct, but \sysname{T_{<:}} is a typed
language; he wants to take advantage of this to build his program
incrementally using the teacher's provided incremental verifier. As a
first attempt, he wants to verify $P_1$:
$$
\recb{\s\z}{\s\z} x y {\s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$, and
he could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \\&\qquad
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program. Derivation $\md_2$
used to live in the empty context, we are then free to use it in the
new context where $\var{add}:\cst{nat}\to\cst{nat}$, because System
\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
with the derivation $\finfer{\var u}$ inferred recursively for the new
recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{odd}$, but $\var{add}$ awaits $\cst{nat}$ arguments. When the
system will generate this derivation, it will insert a coercion
$\cst{odd} \leq \cst{nat}$ at these two points. Taking advantage of
$\md_2$ and $\md_4$ the system succeeds with a new derivation made out
of these, without having to regenerate them because it knows that
$\finfer{\fget\md} = \md$.

@@

\begin{mathleft}
  \quad\case{\app{M}N} \\
  \quad\quad\letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
  \quad\quad\letd {\pair {A_2} {\md_2}} {\finfer{N}} \\
  \quad\quad\letd {\md_{\leq}} {\fleq {A_1} {A_2}} \\
  \quad\quad
  \match{\md_{\leq}} \\
  \quad\quad\quad\case{\infer{ }{\vdash A\leq A}}
  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}} \\
  \quad\quad\quad\case{\_}
  \pair {B} {\infer{\md_1 \and \infer*{\md_2 \and \md_{\leq}}{\vdash N
      : A_1}}{\vdash \app M N : B}}
\end{mathleft}

\noindent
The $\syntax{let}$ construct is just syntactic sugar for a one-branch
$\syntax{case}$; we use an auxiliary function $$\fleq{}{}\ :\ \prd A
{\cst{tp}} \prd B {\cst{tp}} \vdash A\leq B = \ldots$$ trying to
construct a subtyping derivation for its argument, or failing with an
error. Notice the slight optimization to the size of the derivation:
if the types $A_1$ and $A_2$ are the same syntactically, we don't
generate a subtyping rule. This is done by matching on the returned
derivation $\md_\leq$.

Then comes the abstraction case $\tlam x A M$. The difficulty is that
the recursive call $\finfer{M}$ must be done in an enlarged
environment where $\vdash x : A$, otherwise the check that the
returned open derivation $\md$ is correct would fail ($x$ is
unknown). We thus introduce a new construct $\matchin M \Gamma \ldots$
and its $\syntax{let}$ counterpart ($\match M \ldots$ being syntactic
sugar for $\matchin M \envnil \ldots$). Even then, in the recursive
call, the variable case $\var x$ would fail since we would have no way
of telling its type anymore. Thus we substitute all occurrences of
$\var x$ with the \emph{output} of function $\finfer{}$ on it (since
it is known at this point), \ie\ a pair of $A$ and $\md_1$. For this
substitution to be well-typed, we now need to coerce this output to a
valid input (a term). We introduce an operator which is in some sense
the \emph{inverse} of $\finfer{}$: operator $\fget{}$ maps a pair of a
type and a derivation to the term it types.

@@

At this point, we get back $B$ and $\md$ from the recursive call. We
know that $B$ can't depend on $\md_x$ (because a System
\sysname{T_{<:}} type is always closed), but $\md$ can: it is not
\emph{per se} a hypothetical derivation but an \emph{open} derivation
where a special judgement named $\md_x$ can occur. To return the
complete derivation of $\vdash\tlam A x M : A \to B$, we must then
close, or ``hypothesize'' $\md$ by $\md_x$, and then apply rule
\rulename{Lam}:

@@

Concerning the treatment of environments, the case of $\recb M N x y
P$ is similar: we first compute recursively the derivations $\md_M$
and $\md_N$ which should be well-typed in the current environment, and
then $\md_P$ in the enlarged environment, with variables $\var x$ and
$\var y$ in $P$ substituted with the (future) result of $\finfer{}$.

Remains one problem with subtyping: terms $N : A_N$ and $P : A_P$
should be coercible to the same type $A$, but $P$ should be typed
under $\vdash y:A$, which is not known before typing $P$. A trick to
break the loop is to iterate the typing of $P$ with more and more
general types, until we reach a fixpoint (we reach a fixpoint if there
is a unique more general type); in our particular case of subtyping
relation, this means iterating only twice: we will then attain
\cst{nat}.

@@

The type-checker is now done. There is no case for variables, as we
won't ever meet this case: we already substituted all variables with
their result derivation. Evaluation only need to make sure that in
this case, the derivation will be returned, \ie\ that it reduces the
following way:
$$
\finfer{\fget{\pair A \md}} = \pair A \md
$$
Now $\fget{}$ can be derived automatically from the type of
$\finfer{}$: it is its inverse. It has definition:

% TODO confirmer qu'on met bien la def. des f^0
\begin{mathleft}
  \fget{}\ :\ \prd{\{M\}}{\cst{tm}} \prd {A}{\cst{tp}} \prd{\md}{(\vdash
    M:A)} \cst{tm} =
  \\\qquad\quad \lamd M \lamd A \lamd \md M
\end{mathleft}

@@

The idea of untrustedly generating certificates verified \emph{a
  posteriori} is very wide-spread in \sysname{LCF}-style proof
assistants: during manual proof search, the interaction is driven by
\emph{tactics}: %TODO cite
they are untrusted programs generating pieces of proofs, assembled by
other, higher-order tactics; verification of the generated proofs is
done when the evaluation of all tactics is complete. The differences
with our mechanism are two-fold: first, tactics usually take one (or
several) \emph{goal}(s) as input, \ie\ holes in a proof, whereas our
functions can take any term; secondly, proof-check is usually done at
the end of the evaluation, which can lead to less tractable bugs; by
making evaluation and checking mutually recursive, we choose to detect
errors earlier in these witnesses: exactly at the point where they are
generated.

Introducing the ``inverse'' of the derivation inference function has
two purposes: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special variable
construct $\{x:A\}$ denoting free variables is added to the syntax of
terms, and the rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
% TODO: vérif ce paragraphe après avoir parlé de fcts inv
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would arrive to the unhandled case $\finfer{\var
  x}$. This requires evaluating the type-checker with a carefully
designed strategy (see \ref{sec:comput}).

@@

To write deltas by sharing subterms, we need to be able to address any
subterm and instantiate its free variables. For this, we define the
\emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
contextual metavariables $\meta X$, $\meta Y$\ldots\ as in
\cite{nanevski2008contextual}. Metavariables stand for well-typed
terms defined in a repository, and are instantiated wrt. their free
variables. A repository also contains a distinguished metavariable
corresponding to the head of the term stored in it. For instance, the
repository
$$
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{o}} &:
    \cst{tm}

\end{array}\right), \smeta X {}
$$
is one of the possible way of slicing of the term
$$
\fbox{$
  \app {\cst{lam}} \lam x
  \fbox{$
    \app {\cst{lam}} \lam y
    \fbox{$
      \app {\cst{app}} \app {\var y}
      (\fbox{$
        \app{\cst{app}}\app{\var x}{\cst o}
      $})
    $}
  $}
$}
$$
where all boxed subterms are referable slices.

@@

We distinguish syntactically between \emph{checkable} $M$ and
\emph{inferrable} $F$ objects (resp. \emph{canonical} and
\emph{atomic}), an idea dating back to \cite{pierce2000local}:
canonical objects are checked against a given type, whereas we can
synthesize the type of atomic objects; the coercion from the first
grammatical category to the second is where we verify equality between
the synthesized and the checked types.

@@

@@

Our algorithm accepts exactly the well-typed objects of the usual \LF,
except that we thread a repository $\mr$ and store in it all
well-typed applicative subterms $F$ along with their classifier $A$
and the environment in which they are well-typed.

@@

% \begin{theorem}
%   If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jem\mmr\Gamma M {M'} A$
% \end{theorem}

% \begin{theorem}
%   If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jlfm\Gamma {(\checkoutrr
%     M)} (\checkoutrr A)$
% \end{theorem}

@@

\section{Related work}
% 1-2 pages
